# Getting and Cleaning Data Course Project

******

## Instructions

The purpose of this project is to demonstrate your ability to collect, work with, and clean a data set.

### Review criteria 

1. The submitted data set is tidy.
2. The Github repo contains the required scripts.
3. GitHub contains a code book that modifies and updates the available codebooks with the data to indicate all the variables and summaries calculated, along with units, and any other relevant information.
4. The README that explains the analysis files is clear and understandable.
5. The work submitted for this project is the work of the student who submitted it.

### Description of the problem

The purpose of this project is to demonstrate your ability to collect, work with, and clean a data set. The goal is to prepare tidy data that can be used 
for later analysis. You will be graded by your peers on a series of yes/no 
questions related to the project. You will be required to submit: 

1. a tidy data set as described below, 
2. a link to a Github repository with your script for performing the analysis, and 
3. a code book that describes the variables, the data, and any  transformations or work that you performed to clean up the data  called CodeBook.md. 
        
You should also include a README.md in the repo with your scripts. This repo  explains how all of the scripts work and how they are connected.

One of the most exciting areas in all of data science right now is wearable 
computing - see for example this article . Companies like Fitbit, Nike, and 
Jawbone Up are racing to develop the most advanced algorithms to attract new 
users. The data linked to from the course website represent data collected 
from the accelerometers from the Samsung Galaxy S smartphone. A full 
description is available at the site where the data was obtained:
    
<http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones>

Here are the data for the project:
    
<https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip>

You should create one R script called run_analysis.R that does the following.

1. Merges the training and the test sets to create one data set.
2. Extracts only the measurements on the mean and standard deviation for each measurement.
3. Uses descriptive activity names to name the activities in the data set
4. Appropriately labels the data set with descriptive variable names.
5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

******

### Getting the data

Load the `dplyr` package.
```{r message=FALSE}
library(dplyr)
```

Download the data.
```{r, cache=TRUE}
fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(fileUrl, destfile = './data.zip')
unzip('./data.zip')
```

Refer to `README.txt` for a detailed description of the data.
The folder `UCI HAR Dataset` contains the *test* and *train* data. 

Create a `dplyr` data frame for each subject id in test and train sample.
```{r, cache=TRUE}
subject_test <- tbl_df(read.table('./UCI HAR Dataset/test/subject_test.txt'))
subject_train <- tbl_df(read.table('./UCI HAR Dataset/train/subject_train.txt'))
```

Create a `dplyr` data frame for each features data in test and train sample.
```{r, cache=TRUE}
data_test <- tbl_df(read.table('./UCI HAR Dataset/test/X_test.txt'))
data_train <- tbl_df(read.table('./UCI HAR Dataset/train/X_train.txt'))
```

Create a `dplyr` data frame for each activity in test and train sample.
```{r, cache=TRUE}
activity_test <- tbl_df(read.table('./UCI HAR Dataset/test/Y_test.txt'))
activity_train <- tbl_df(read.table('./UCI HAR Dataset/train/Y_train.txt'))
```

Merge subject, activity and data set for train and test data. This creates three `dplyr` data frames.
```{r, cache=TRUE}
subjects <- rbind(subject_train, subject_test)
activity <- rbind(activity_train, activity_test) 
data <- rbind(data_train, data_test)
```

For convenience, properly name columns for features in 'data' data frame. 
First load the features data and activity _names_ from the `features.txt` and `activity_labels.txt` files. 
```{r, cache=TRUE}
features <- read.table('./UCI HAR Dataset/features.txt')
labels <- read.table('./UCI HAR Dataset/activity_labels.txt')
```
Use the `rename` function to rename the variables names of the `subjects` and `activity` data frames defined above.
```{r, cache=TRUE}
subjects <- rename(subjects, subject = V1)
activity <- rename(activity, activity_num = V1)
colnames(data) <- features$V2
```

Merges the training and the test sets to create one data set named `df`.
```{r, cache=TRUE}
df <- tbl_df(cbind(subjects, activity, data))
```
Preview the _untidy_ data frame.
```{r, cache=TRUE}
df[1:10, 1:4]
```

******

### Tidy the data

Extract mean and standard deviation measurements. Use the `grep` function to search for the words _mean_ and _standard deviation_ in the features columns of `df`.
```{r, cache=TRUE}
mean_std_features <- grep(".[Mm]ean.|.[Ss]td.",names(df))
df<- subset(df,select=c(1,2,mean_std_features))
```
Preview the names of the first few features:
```{r, cache=TRUE}
head(names(df), n=20L)
```

Use descriptive activity names for activity data frame. This requires to merge the `df` data frame and the `labels` data frame by the *activity number*. 
```{r, cache=TRUE}
names(labels) <- c('activity_num', 'activity')
df <- merge(labels, df , by='activity_num', all=TRUE); rm(labels)
```
A new column is created. We only need the column with the activity label and not the activity number. 
Use the functions `select` and `arange`

Rearragnge columns to show subject first and activity second.
```{r, cache=TRUE}
df <- df %>%
    select(3, 2, 4:89) %>%
    arrange(subject)
```

Use descriptive names for mean and standard devaition features.
According to the lectures, good habits include:

*   All lower case when possible.
*   Descriptive (_Diagnosis_ versus _Dx_).
*   Not duplicated.
*   Not have underscores or dots or white spaces.
*   Variables with chacacter values should be made into factor variables (depending on context).

```{r, cache=TRUE}
names(df) <- tolower(names(df))
names(df) <- gsub('acc', 'acceleration', names(df))
names(df) <- gsub('gyro', 'angularvelocity', names(df))
names(df) <- gsub('mag', 'magnitude', names(df))
names(df) <- gsub('bodybody', 'body', names(df))
names(df) <- gsub('freq', 'frequency', names(df))
names(df) <- gsub('anglet', 'angletime', names(df))
names(df) <- gsub('meangravity', 'gravity', names(df))
names(df) <- gsub('^t|\\(t', 'time', names(df))
names(df) <- gsub('^f', 'frequency', names(df))
names(df) <- gsub('-|\\(|\\)|,', '', names(df))
names(df) <- gsub('anglexgravitymean', 'anglegravitymeanx', names(df))
names(df) <- gsub('angleygravitymean', 'anglegravitymeany', names(df))
names(df) <- gsub('anglezgravitymean', 'anglegravitymeanz', names(df))
```

Let us preview the first and last few variables names:
```{r, cache=TRUE}
head(names(df))
tail(names(df))
```
The _tidied_ data is now:
```{r, cache=TRUE}
df[1:10, 1:4]
```
The structured of our _tidied_ data is:
```{r, cache=TRUE}
str(df)
```
_I believe a possible improvement would be to group the data so that we have columns denoting, the subject, the activity, the time domain, frequency and angle gravity._

******

### Second tidy data

From the data set `df`, create a second, independent tidy data set with the average of each variable for each activity and each subject.
For this, we use the `aggregate` function and arrange according to subject and activity with `arrange`.
```{r, cache=TRUE}
tidySet <- aggregate(.~subject + activity, df, mean)
tidySet <- tidySet %>% arrange(subject, activity)
```
